Disentangling linkages between satellite-derived forest structure and productivity essential biodiversity variables.

### 2023-07-24 scripts/manuscript/manuscript.qmd
sent to meg and tx 2023-07-21b

---
title: Disentangling linkages between satellite derived forest structure and productivity essential biodiversity variables.
author:
  - name: Evan R. Muise
    email: evan.muise@student.ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
    attributes:
        corresponding: true
  - name: Margaret E. Andrew
    email: M.Andrew@murdoch.edu.au
    affiliations: 
        - id: murd
          name: Murdoch University
          department: Environmental and Conservation Sciences and Harry Butler Institute
          city: Murdoch, WA, Australia
          postal-code: 6150
  - name: Nicholas C. Coops
    email: nicholas.coops@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Txomin Hermosilla
    email: txomin.hermosillagomez@NRCan-RNCan.gc.ca
    affiliations: 
        - id: cfs
          name: Natural Resources Canada
          department: Canada Forest Service (Pacific Forestry Centre)
          address: 506 Burnside Rd W
          city: Victoria, BC, Canada
          postal-code: V8Z 1M5
  - name: A. Cole Burton
    email: cole.burton@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Stephen S. Ban
    email: Stephen.Ban@gov.bc.ca
    affiliations: 
        - id: bcp
          name: Ministry of Environment and Climate Change Strategy
          department: BC Parks
          address: 525 Superior Street
          city: Victoria, BC, Canada
          postal-code: V8V 1T7
abstract: Under a changing climate and in light of the biodiversity crises, monitoring change in biodiversity at broad scales has become ever more imperative. Key to regular biodiversity monitoring over time and space is the definition and selection of variables capable of accurately monitoring biodiversity trends. One framework is the Essential Biodiversity Variables (EBV), which are designed to be analogous to the Essential Climate variables and capable of monitoring biodiversity globally while also being complementary to one another within EBV classes. Five of six EBV classes are well suited to being monitored using satellite remote sensing, with particular focus warranted on the ecosystem structure and function classes. In this paper, we seek to address the complementarity of variables describing forest structure, as imputed across the entirety of British Columbia, Canada, with variables representing ecosystem functioning, namely the Dynamic Habitat Indices, a yearly summary of productivity indices. Both sets of variables have previously been linked with biodiversity metrics across a range of scales. Using redundancy analysis, we find that forest structural attributes and the DHIs are essentially decoupled from each other, with the forest structure variables explaining 14% of the variation in the DHIs. Further, we explore how the proportion of variance explained in the DHIs varies with either primary structure variables, such as canopy height and cover, vs more complex modelled forest structural attributes, such as aboveground biomass and volume. Results suggest there is large overlap between the two types of forest structure variables suggesting little added benefit of derived attributes beyond height and cover, in this context. Overall, we find that forest structure as an ecosystem structure EBV and the DHIs as an ecosystem function EBV represent markedly different processes as observed by remote sensing and they are highly complementary across large environmental gradients, suggesting that in an earth-observation based monitoring program both are needed.
keywords: 
  - remote sensing
  - landsat
  - forest structure
  - forest productivity
  - dynamic habitat indices
  - essential biodiversity variables
  - redundancy analysis
date: last-modified
bibliography: [bibliography.bib, packages.bib]
csl: remote-sensing-of-environment.csl
format:
  elsevier-pdf:
    keep-tex: true
    keep-md: true
    journal:
      name: Remote Sensing of Environment
      formatting: review
      model: 3p
      cite-style: authoryear
      layout: twocolumn
  docx: 
    reference-doc: my-styles.docx
    keep-md: true
  pdf: 
    pdf-engine: xelatex
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: false
fig-width: 300
---

@@

@@

With biodiversity in decline, and facing extinction rates above the
background extinction rate [@thomas2004; @urban2015], as well as the
homogenization of communities at various scales [@mcgill2015], it is
integral to be able to monitor how biodiversity is changing across the
globe. In response, the global biodiversity community is making efforts
to assess and halt the degradation of biodiversity. The Group on Earth
Observations Biodiversity Observation Network has developed the
Essential Biodiversity Variables [EBVs, @pereira2013], designed as an
analog to the Essential Climate Variables framework [@bojinski2014].
EBVs are designed to be global in scope, relevant to biodiversity
information, feasible to use, and complementary to one another
[@skidmore2021]. While it can be difficult, time consuming, and
expensive to collect data on biodiversity across wide swaths of land and
varying ecosystems, EBVs, which can be correlated to sampled
biodiversity information, allow for the monitoring and assessment of
protected area effectiveness and ecosystem health at large spatial
scales [@hansen2021]. There are six EBV classes, each of which
correspond to a different facet of biodiversity: species populations,
species traits, community composition, ecosystem structure, ecosystem
function, and genetic composition [@pereira2013].

Satellite remote sensing has proven to be capable of measuring five of
the six EBV classes, the exception being genetic composition, which
requires in-situ observation and sampling [@skidmore2021]. Species
populations, and in turn community composition, can be assessed with
very-high-resolution imagery to identify tree species at the tree-crown
scale [@fassnacht2016; @graves2016]. Additionally, species traits such
as vegetation phenology have been observed at the single-tree scale
using, for example, PlanetScope imagery and drone-based measurements
[@wu2021]. However, the spatially limited and often *ad-hoc* data
collection approaches associated with monitoring individuals is not
conducive to the global or regional scales required for biodiversity
trend assessment [@valdez2023], in addition to the difficulties
associated with extending species-level analyses to broader extents
[@fassnacht2016; @graves2016].

The two landscape-level EBVs (ecosystem structure and function) are well
suited to be examined at large spatial scales using coarser spatial
measurements, such as those taken from satellites by the Moderate
Resolution Imaging Spectroradiometer [MODIS, @zhang2003], the Landsat
imaging systems [@fisher2006], or Sentinel-2 [@helfenstein2022;
@darvishzadeh2019] programs. These mid-resolution satellites can monitor
processes at broader extents, but their coarse spatial resolution
removes the ability to relate these traits to individual organisms. As a
result, satellite remote sensing data has been shown to be arguably the
most effective at monitoring ecosystem based EBVs focused on structure
and function. These EBVs classes can be monitored at regional to global
extents through the use of optical imagery [@cohen2004], as well as
active sensors such as lidar (light detection and ranging) and radar
[@guo2017; @lefsky2002; @lang2021; @neuenschwander2019; @coops2016].

## EBV - Ecosystem Structure

Forest structural diversity has been linked to biodiversity at various
scales [@guo2017; @bergen2009; @gao2014]. Structural attributes range in
complexity from simple (canopy cover; canopy height), to more complex
(vertical and horizontal structural complexity) to modelled (aboveground
biomass; basal area), all of which can be assessed using lidar data
[@coops2021]. A suite of these lidar-derived attributes have been used
as local indicators of biodiversity, including simple metrics such as
canopy cover and canopy height as well as derived metrics including
vertical profiles, aboveground biomass, or lidar derived indices
[@lefsky1999; @guo2017; @coops2016]. Other second order derived metrics
such as canopy texture, height class distribution, edges, and patch
metrics have also been used to examine habitat and biodiversity at
landscape scales [@bergen2009]. Advances in satellite remote sensing
processing have now allowed 3D forest structure data to be imputed
across wide spatial scales [@matasci2018; @coops2021] using data fusion
approaches involving collected lidar data and optical/radar data.

Increased forest structural complexity has been hypothesized to create
additional niches, leading to increased species diversity [@bergen2009],
which has been frequently demonstrated using avian species diversity
metrics [@macarthur1961]. For example, @herniman2020 used spectral and
lidar derived forest structure data to model avian habitat suitability;
@clawges2008 found that lidar derived forest structural attributes are
capable of identifying habitat types associated with avian species in
pine/aspen forests; @goetz2007 used canopy structural diversity to
predict bird species richness, finding that canopy vertical distribution
was the strongest predictor of species richness. Forest structural
metrics have also been used to study biodiversity in other clades as
well [@davies2014; @nelson2005].

## EBV - Ecosystem Function

With respect to ecosystem function, energy availability in an ecosystem
has shown to be a predictor of species richness and abundances at
various scales [@chase2002; @radeloff2019; @coops2019; @razenkova2023],
and is measurable using satellite remote sensing via the use of various
vegetation indices [@huete2002; @radeloff2019]. Vegetation indices,
which are indicative of photosynthetic activity, are commonly used as
proxies of gross primary productivity [@huang2019]. These vegetation
indices have also been used to assess patterns in biodiversity at single
time points [@bonn2004], and more recently, through yearly summaries of
productivity [@berry2007; @radeloff2019]. The relationship between
energy availability and biodiversity occurs via various hypothesized
mechanisms, such as the available energy hypothesis [@currie2004;
@wright1983], the environmental stress hypothesis [@currie2004], and the
environmental stability hypothesis [@williams2008]. These three
hypotheses have in turn been linked to patterns of annual surface
reflectance in remote sensing data [@berry2007; @radeloff2019].

@berry2007 first explored this idea by proposing the linkage of
intra-annual summaries of MODIS-derived GPP to dispersive bird species.
This idea was further refined into the Dynamic Habitat Indices [DHIs,
@coops2008], which have now been shown to be well suited to assess the
three aforementioned hypotheses at global scales [@radeloff2019]. The
cumulative DHI calculates the total amount of energy available in a
given pixel over the course of a year. Cumulative DHI is strongly linked
to the available energy hypothesis, which suggests that with greater
available energy species richness will increase [@wright1983]. The
minimum DHI, which calculates the lowest productivity over the course of
a year can be matched to the environmental stress hypothesis, which
proposes that higher levels of minimum available energy will lead to
higher species richness [@currie2004]. Finally, the variation DHI, which
calculates the coefficient of variance in a vegetation index through the
course of a year, corresponds to the environmental stability hypothesis
which states that lower energy variation throughout a year will lead to
increased species richness [@williams2008].

## Biodiversity Monitoring with EBVs

Biodiversity monitoring programs often require a range of information in
order to accurately assess changes in ecological integrity
[@lindenmayer2010]. Choosing datasets that are most closely related to
the phenomenon of interest in a given application allows for direct
connections to monitoring results and management actions [@pressey2021].
With the advent of large-extent monitoring methods like satellite remote
sensing, and a proliferation of potential EBVs datasets, it becomes
important to assess the interrelationships between these datasets, and
assess their complementary of the information to reduce the duplication
of efforts [@pereira2013; @skidmore2021]. When strong relationships are
present between EBVs, it is possible to assess the ecological
relationships between potential EBVs. On the other hand, when datasets
do not appear related, they may be well suited to be used in monitoring
programs together, as complementary EBVs.

Linkages between forest ecosystem structure and function have been
examined within a remote sensing context for over 20 years [@huete2002,
@knyazikhin1998; @myneni1994]. While there is significant theoretical
and empirical evidence for their relationship at single time points
(e.g. within a single image) [@myneni1994], various relationship
directions and shapes have been found between forest structure and
function metrics [@ali2019]. Hypothesized mechanisms such as niche
complementarity have shown that aboveground biomass increases with stand
structure [@zhang2012], while asymmetric competition for light can
reduce forest productivity with increased structural complexity
[@bourdier2016]. The relationship in particular between forest
structural diversity metrics - which are now more accurately and
comprehensively derived from lidar data - and temporal variation in
functional metrics, specifically the metrics of ecosystem productivity
via the DHI framework, have yet to be fully examined.

The overall goal of this paper is to assess patterns of forest ecosystem
structure and function, as observed using remote sensing technologies,
and their complementarity across a wide range of ecosystems encompassing
significant environmental gradients. To do so, we synthesize data from
moderate-scale remote-sensing derived metrics of forest structure,
represented as both simple ALS-extracted metrics of canopy height, cover
and vertical complexity, as well as modelled forest structure attributes
including volume and aboveground biomass, with a well-established remote
sensing derived index on ecosystem function. First, we examine how
ecosystem structure and function complement one another across a large
environmental gradient and then compare the simple and modelled
representations of forest structure to different levels of ecosystem
function. This question is important as it provides insights to the EBV
community around complementarity of remote sensing metrics when
describing the structure and function of ecosystems and proposes a
method to examine potential overlap when generating remote sensing EBVs.

Our second question examines the independent and shared relationships of
lidar derived forest structural attributes, with modelled forest
structure, on ecosystem function. This provides insight into the
selection of remote sensing attributes to use when developing EBVs
within a single EBV class. Remote sensing datasets can comprise
relatively unprocessed observations, in this case lidar-imputed measures
of height and cover vs modelled attributes, such as biomass and volume,
which involve the use the statistical relationships with field data to
transform the observations into more refined data products. Assessing
which of these two (or combination of the two) approaches has stronger
or weaker relationships with estimates of ecosystem function provides
insights into the choice of data used to build EBVs. Lastly, we examine
how the primary and modelled structure variables partition the variance
of the DHIs within key biomes and forest types across a large
environmental range, examining to what extent ecosystem and forest types
impacts these relationships and thus providing insight into the
applicability of these results globally.

# Methods

@@

British Columbia is the westernmost province of Canada, and is home to a
variety of terrestrial ecosystems. Approximately 64% of the province is
forested, with large environmental and topographic gradients
[@pojar1987; @bcministryofforests2003]. The Biogeoclimatic Ecosystem
Classification (BEC) system identifies 16 zones based on the dominant
tree species and the ecosystems general climate (@fig-study-map).

```{r}
#| label: fig-study-map
#| fig-cap: Location of biogeoclimatic ecosystem classification (BEC) zones in British Columbia.

@@

These zones can be further split into subzones, variants, and phases
based on microclimate, precipitation, and topography [@pojar1987]. To
examine trends across the large environmental gradients, we group the
BEC zones into five broad biomes: the southern interior, northern
interior, montane, alpine, and coastal groups similar to @hamann2006. We
also report each BEC zone's average climate data, according to @wang2020
(@tbl-bec-group).

```{r}
#| label: tbl-bec-group
#| tbl-cap: BEC Zones, their aggregate groups, and their average climate values for precipitation, maximum temperature, and minimum temperature. Climate data from Wang et al. 2016.

# bec_tab %>%
#   gt() %>%
#   cols_label(
#     group = "BEC Group",
#     zone = "BEC Zone",
#     zone_nm = "Full Name",
#     ppt = "Precipitation (mm/yr)",
#     tmax = "Max Temperature (°C)",
#     tmin = "Min Temperature (°C)"
#   ) %>%
#   as_word()

knitr::kable(
  bec_tab,
  col.names = c(
    "BEC Group",
    "BEC Zone",
    "Full Name",
    "Precipitation (mm/yr)",
    "Max Temperature (°C)",
    "Min Temperature (°C)"
  )
)
```

@@

### Forest Structure

We used a suite of forest structure variables (canopy height, canopy
cover, structural complexity \[coefficient of variation of height
returns\], basal area, aboveground biomass, and gross stem volume). This
dataset was created for the year 2015 at a 30 m spatial resolution
according to @matasci2018. In brief, the method used a set of lidar
collections and field plots across Canada, and imputed the remaining
pixels using a random forest k-Nearest Neighbour approach on
Landsat-derived surface reflectance and auxiliary data such as
topography. Detailed information on the creation of this dataset can be
found in @matasci2018.

@@

We use an established set of indices of annual productivity shown to be
related to global biodiversity trends: the Dynamic Habitat Indices
[@radeloff2019]. The DHIs are a set of satellite remote sensing derived
productivity variables that summarize the cumulative amount of available
energy, the minimum available energy, and the variation in available
energy throughout a given year [@berry2007; @radeloff2019]. The DHIs
have previously been produced at a global extent using MODIS imagery,
and have previously been used to assess alpha [@radeloff2019] and beta
[@andrew2012] diversity, species abundances [@razenkova2023], and
construct novel ecoregionalizations [@coops2009; @andrew2013]. Recent
studies have began to examine how these indices can be constructed at a
finer spatial resolution by using multi-annual Landsat imagery to
generate a single synthetic year of monthly observations
[@razenkova2022].

The DHIs were calculated according to [@razenkova2022, Razenkova et al.,
In Press] for all of terrestrial British Columbia. In brief, Google
Earth Engine [@gorelick2017] was used to obtain all valid Landsat pixels
for a given study area, filtering out pixels containing shadows, clouds,
and cloud shadows within each image using the fmask algorithm
[@zhu2012], then calculated the NDVI for each pixel in each image. They
then calculated the median NDVI value for each month across a ten year
time span (2011-2020) to generate a synthetic year of monthly data. The
sum, minimum, and coefficient of variation across this synthetic year of
NDVI values was then calculated. More detailed information can be found
in Razenkova et al. (In Press).

@@

# Sampling

>>>>

---
title: Disentangling linkages between satellite-derived forest structure and productivity essential biodiversity variables.
author:
  - name: Evan R. Muise
    email: evan.muise@student.ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
    attributes:
        corresponding: true
  - name: Margaret E. Andrew
    email: M.Andrew@murdoch.edu.au
    affiliations: 
        - id: murd
          name: Murdoch University
          department: Environmental and Conservation Sciences and Harry Butler Institute
          city: Murdoch, WA, Australia
          postal-code: 6150
  - name: Nicholas C. Coops
    email: nicholas.coops@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Txomin Hermosilla
    email: txomin.hermosillagomez@NRCan-RNCan.gc.ca
    affiliations: 
        - id: cfs
          name: Natural Resources Canada
          department: Canada Forest Service (Pacific Forestry Centre)
          address: 506 Burnside Rd W
          city: Victoria, BC, Canada
          postal-code: V8Z 1M5
  - name: A. Cole Burton
    email: cole.burton@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Stephen S. Ban
    email: Stephen.Ban@gov.bc.ca
    affiliations: 
        - id: bcp
          name: Ministry of Environment and Climate Change Strategy
          department: BC Parks
          address: 525 Superior Street
          city: Victoria, BC, Canada
          postal-code: V8V 1T7
keywords: 
  - remote sensing
  - landsat
  - forest structure
  - forest productivity
  - dynamic habitat indices (DHIs)
  - essential biodiversity variables
  - redundancy analysis (RDA)
date: last-modified
bibliography: [bibliography.bib, packages.bib]
csl: remote-sensing-of-environment.csl
format:
  elsevier-pdf:
    keep-tex: true
    keep-md: true
    journal:
      name: Nature Scientific Reports
      formatting: review
      model: 3p
      cite-style: authoryear
      layout: twocolumn
  docx: 
    reference-doc: my-styles.docx
    keep-md: true
  pdf: 
    pdf-engine: xelatex
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: false
fig-width: 300
number-sections: true
---

@@

**Abstract**

Key to regular biodiversity monitoring is the definition and selection
of variables capable of accurately monitoring biodiversity trends. One
framework is the Essential Biodiversity Variables (EBV), which are
standardized, comparable data that represents a minimum set of
biological information required for monitoring biodiversity change at
large spatial extents, with each class being complementary to one
another -- i.e., capturing unique biological information. The
interrelationships between forest ecosystem structure and function and
the complementarity between these two classes of EBVs across a range of
ecosystems encompassing significant environmental gradients are examined
using redundancy analysis (RDA) and variation partitioning. We found
that the primary and modelled forest structure variables are not
strongly related to the Dynamic Habitat Indices (DHIs), with the
structural information only explaining 16% of the variation in the DHIs,
indicating that structure and function EBVs are suitable to be used
together for monitoring forest ecosystems. We also found that forest
types and ecosystems dominated by coniferous trees had less variation
explained, potentially due to the temporal decoupling of canopy cover
and the DHIs. We suggest that biodiversity researchers focus on using
forest structural attributes derived directly from lidar data alongside
the DHIs as complementary EBVs for monitoring forest ecological
integrity.

@@

Monitoring the changing state of biodiversity globally is key to
understanding and mitigating the accelerated extinction risk for many
species [@thomas2004; @urban2015] and the homogenization of biotic
communities at various spatial and temporal scales
[@mcgillFifteenFormsBiodiversity2015]. The Group on Earth Observations
Biodiversity Observation Network has developed the Essential
Biodiversity Variables [EBVs, @pereira2013], which are standardized,
comparable data that represents a minimum set of biological information
required for monitoring biodiversity change at large spatial extents
[@skidmore2021]. The EBVs are designed as an analog to the Essential
Climate Variables framework [@bojinski2014] to be global in scope,
relevant to biodiversity information, feasible to implement and use, and
complementary to one another [@skidmore2021]. While it can be difficult,
time consuming, and expensive to collect data on biodiversity across
wide swaths of land, EBVs can be combined with sampled biodiversity
information to allow for the spatial extrapolation of more sparsely
collected in situ data [@cavender-bares2022]. The standardization and
global availability of the EBVs allows for the assessment and monitoring
of protected area effectiveness and ecosystem health at large spatial
scales [@hansen2021], which can then be integrated into policy
development [@geijzendorffer2016]. There are six EBV classes, each of
which provide information on a different facet of biodiversity: species
populations, species traits, community composition, ecosystem structure,
ecosystem function, and genetic composition [@pereira2013]. While these
EBV classes can be related to one another under a variety of ecological
and evolutionary mechanisms, they provide distinct information; multiple
in concert can provide a more complete picture on the status and trends
of biodiversity in a region.

The global coverage and moderate spatial resolutions of many optical
satellite remote sensing missions [@wulder2022], combined with open-data
policies [@wulder2012], make these datasets exceptionally well suited
for contributing to some EBV classes [@skidmore2021]. The six EBV
classes can be divided into two general approaches: species-focused and
ecosystem-focused [@fernández2020]. Genetic composition, species
populations, and species traits are species-focused EBVs, with community
composition being directly related to species populations. While
species-focused and community composition EBVs can be assessed from
measurements at the tree-crown scale using very-high spatial resolution
imagery [@fassnacht2016; @graves2016; @wu2021], the spatially
limited--and often *ad-hoc*--data collection associated with monitoring
individuals is not conducive to the global or regional scales often
required for biodiversity trend assessments [@valdez2023]. In contrast,
two ecosystem-based EBV classes (ecosystem structure and ecosystem
function) are well suited to be monitored across large extents using
moderate spatial resolution optical imagery, such as the Moderate
Resolution Imaging Spectroradiometer [MODIS, @zhang2003], Landsat
imaging systems [@wulder2022] or Sentinel-2 [@helfenstein2022;
@drusch2012] programs, especially in forested environments
[@pettorelli2018; @coops2016].

Ecosystem structure is the composition, abundance, and spatial
arrangement of different ecosystem components, and is indicative of
habitat quality and connectivity [@noss1990; @valbuena2020]. There are
both horizontal and vertical components of ecosystem structure. In
forest environments, forest structural diversity has been linked to
species richness at various spatial scales, ranging from individual
plots to landscapes [@bergen2009; @gao2014]. Structural attributes range
in complexity from simple (e.g., canopy cover, canopy height), to more
complex (e.g., foliage height diversity, leaf area index) to modelled
(e.g., aboveground biomass, basal area), which can be derived using
lidar (light detection and ranging) data [@coops2021]. A suite of these,
and other, structural attributes has been used as indicators of
biodiversity at local to landscape scales [@lefsky1999; @guo2017;
@coops2016; @bergen2009]. Increased forest structural complexity has
been hypothesized to create additional niches, leading to increased
species diversity [@bergen2009], which has been frequently demonstrated
using avian species diversity metrics [@macarthur1961], with lidar
[@clawges2008], a combination of lidar and satellite imagery
[@herniman2020], with some research indicating that canopy vertical
distribution was the strongest predictor of species richness
[@goetz2007]. Although lidar datasets have traditionally been local in
extent, advances in satellite remote sensing processing have allowed the
extension of three-dimensional forest structure attributes nationally in
a wall-to-wall fashion. The wall-to-wall data often uses data fusion
approaches involving lidar data and medium spatial resolution optical
imagery [@matasci2018; @matasci2018b; @coops2021].

Ecosystem functions are measures of ecosystem performance that are the
consequence of one or multiple ecosystem processes [@pettorelli2018;
@ecosyste2005]. Within the EBV framework, vegetation phenology and
metrics of primary productivity are commonly used as potential EBVs
[@skidmore2021]. Energy availability is the amount of accessible and
utilizable energy available for growth in an ecosystem, and is often
measured as net primary productivity [@bonn2004]. Energy availability
has shown to be a predictor of species richness and abundances at
various scales [@chase2002; @radeloff2019; @coops2019; @razenkova2023],
and is measurable using satellite remote sensing via various vegetation
indices [@huete2002; @radeloff2019]. Vegetation indices are indicative
of photosynthetic activity, and are commonly used as proxies of gross
primary productivity [@huang2019]. These vegetation indices enable the
assessment of biodiversity patterns at single time points [@bonn2004],
or through time when using time series data [@berry2007; @radeloff2019].

The relationship between energy availability and biodiversity can be
explained through various hypothesized mechanisms: including available
energy, environmental stress, and environmental stability. These three
hypotheses have been incorporated into the Dynamic Habitat Indices
(DHIs), which summarizes intra-annual dynamics in remotely-sensed
productivity metrics and have been evaluated as broad-scale biodiversity
indicators [@berry2007; @coops2008; @radeloff2019]. According to the
available energy hypothesis [@wright1983] an increase in available
energy results in higher species richness. In the context of the DHIs,
the total energy available within a specific area is quantified using
the cumulative DHI, which sums each productivity observation throughout
the course of a year. The environmental stress hypothesis [@currie2004]
suggests that higher levels of minimum available energy contribute to
greater species richness. The minimum DHI, derived from the lowest
productivity observed over the course of a year, indicates the level of
available energy during stressful periods. The environmental stability
hypothesis [@williams2008] proposes that lower energy variation
throughout the year leads to increased species richness. The variation
DHI captures the coefficient of variation in a vegetation index over the
course of a year, reflecting the stability of energy availability. The
DHIs have previously been produced at a global extent using MODIS
imagery and have been used to assess alpha [@radeloff2019] and beta
[@andrew2012] diversity, species abundances [@razenkova2023], and
construct novel ecoregionalizations [@coops2009; @andrew2013]. Recent
studies have increased the spatial resolution of these indices by
calculating the DHIs from multi-annual Landsat imagery
[@razenkovaMediumresolutionDynamicHabitatInPress].

Choosing datasets that are most closely related to the phenomenon of
interest in a given application allows for direct integration of
monitoring results with management actions [@pressey2021]. With the
advent of large-extent monitoring methods like satellite remote sensing
which have led to a proliferation of potential EBV datasets, it becomes
important to assess the interrelationships between these datasets and
the complementary of the information they provide to reduce redundancy
in future EBV development efforts [@pereira2013; @skidmore2021].
Identifying and exploring strong relationships between EBVs may help
elucidate the ecological relationships between these facets of
biodiversity, such as ecosystem structure and function, at local to
global scales. On the other hand, when variables are not related, they
may be well suited to be used in monitoring programs together, as
complementary EBVs.

Linkages between forest ecosystem structure and function have been
examined within a remote sensing context for decades [@huete2002,
@knyazikhin1998; @myneni1994]. While there is significant theoretical
and empirical evidence for their relationship [@myneni1994], the
direction and significance has not been consistent, with positive,
negative, and non-significant relationships being found [@ali2019].
Hypothesized mechanisms such as niche complementarity suggest that
productivity increases with stand structural complexity [@zhang2012],
while asymmetric competition for light can reduce forest productivity
with increased structural complexity [@bourdier2016]. Some aspects of
forest structure, such as canopy cover and leaf area index, also
directly influence vegetation indices, which are used to estimate
productivity, especially in very dense or very sparse forests
[@pettorelli2005]. Further, the relationship between forest structural
diversity metrics--which are now more accurately and comprehensively
derived from lidar data--and the temporal variation in functional
metrics, specifically the metrics of ecosystem productivity via the DHIs
framework, have yet to be fully examined.

The goal of this study is to examine relationships between forest
ecosystem structure and function, and the complementarity between these
two classes of EBVs across significant environmental gradients. We
synthesize data from moderate-scale remote sensing-derived metrics of
forest structure (simple lidar-extracted metrics of canopy height,
canopy cover, and structural complexity), as well as modelled forest
structure attributes (gross stem volume, basal area, and aboveground
biomass), with a well established remote sensing-derived index for
ecosystem function (DHIs). First, we analyze remote sensing metric
complementarity by examining the multivariate relationship between
ecosystem structure and function and propose a method to examine
potential overlap in remote sensing EBVs. Second, we propose a method
for the selection of remote sensing attributes for developing and
applying EBVs within a single EBV using the independent and shared
relationships of lidar-derived forest structural attributes and modelled
forest structure on ecosystem function. Third, we examine the global
applicability of these results using the relationship between simple and
modelled forest structural attributes with DHIs across a range of
environment gradients and forest types.

# Methods and Materials

@@

British Columbia (BC) is the westernmost province of Canada, and is home
to a variety of terrestrial ecosystems. Located between the Pacific
Ocean and the Rocky Mountains, approximately 64% of the province is
forested, with large environmental and topographic gradients
[@pojar1987; @bcministryofforests2003]. The majority of the forests of
BC are coniferous forests (87.4%), with the remaining forests being
broadleaf (8.9%), wetland-treed (2.7%) and mixed-wood (0.9%)
[@hermosilla2022]. There is a large topographic gradient in the
province, with elevations ranging from sea level to over 4000 m, with
much of the province being mountainous (elevations above 1000 m)
[@valentine1978soil]. Wildfires are a regular disturbance in the
province, with the historical fire regime being disrupted due to
management actions, leading to larger, higher intensity fires becoming
more common in recent years [@brookes2021]. In addition, forestry is a
large industry in the province, with approximately 136000 ha of the
province harvested accounting for \~54 million cubic metres of volume of
wood in 2021 [@canadiancouncilofforestministers2023].

BC is stratified into 16 zones based on the dominant tree species and
climate by the Biogeoclimatic Ecosystem Classification (BEC) system
(@fig-study-map). To examine trends across the large environmental
gradients, we group the BEC zones into five broad biomes: the Nouthern
interior, Sorthern interior, Montane, Alpine, and Coastal groups,
similar to @hamann2006. We also report the average climate data for BEC
zones from 1991-2020, according to @wang2016 (Table 1).

```{r}
#| label: fig-study-map
#| fig-cap: Location of Biogeoclimatic Ecosystem Classification (BEC) zones in British Columbia.

@@

```{r}
#| warning: false
#| echo: false

bec_tab %>%
  gt() %>%
  cols_label(
    group = "BEC Group",
    zone = "BEC Zone",
    zone_nm = "Full Name",
    ppt = "Precipitation (mm/yr)",
    tmax = "Max Temperature (°C)",
    tmin = "Min Temperature (°C)"
  ) %>%
  tab_header(title = "Biogeoclimatic Ecosystem Classification (BEC) Zones, their aggregated biomes, and their average climate values for precipitation, maximum temperature, and minimum temperature. Climate data from Wang et al. 2016 averaged for 1991-2020.")
```

@@

### National Terrestrial Ecosystem Monitoring System (NTEMS)

#### Best-Available-Pixel

The forest structure, forest disturbance, and land cover layers are
based on a 30-m best-available-pixel (BAP) composite generated by
@hermosilla2016 using the Composite2Change approach. The composites were
developed for each year from 1984-2019 by selecting the best available
imagery from the available growing season Landsat observations, removing
pixels with clouds, cloud shadows, and haze. The methodology uses the
scoring method from @white2014 to select the best pixel for a given year
from Landsat-5 Thematic Mapper, Landsat-7 Enhanced Thematic Mapper Plus,
and Landsat-8 Operational Land Imager imagery. To reduce noise in the
BAP between years, a spectral trend analysis on the Normalized Burn
Ratio was conducted on each pixel, which removes unscreened noise,
detects changes, and temporally interpolates data gaps. This resulted in
a gap-free surface-reflectance composite across the entirety of Canada
[@hermosilla2015]. We utilize data from 2015 for all NTEMS data, except
forest disturbances, where we use all available years of data in the
terrestrial area of BC, Canada.

#### Forest Disturbances

We used a forest disturbance layer generated by @hermosilla2015a to
restrict the analysis to stands older than 35 years as productivity and
structure metrics have been shown to be strongly decoupled in early
regenerative stands [@bolton2017], and also excluding recently
anthropogenically disturbed stands. Changes were detected using spectral
trend analysis and were attributed using an object-based analysis
approach to a disturbance agent (i.e., wildfire, harvest) using a Random
Forests classification model [@hermosilla2015a]. Changes were detected
with an overall accuracy of 89%, attributed with an accuracy of 92%, and
assigned to the correct year with an accuracy of 89% [@hermosilla2016].

#### Land Cover {#sec-lcc}

We used a land cover map generated by @hermosilla2022 to stratify our
samples alongside the BEC zones. The 30-m spatial resolution land cover
map for Canada's forested ecosystems was generated using the Virtual
Land Cover Engine framework [@hermosilla2022]. This framework uses a
Random Forests based classification approach, and integrates logical
land cover transitions using a Hidden Markov Model to generate 12 land
cover classes. Training samples were locally selected using an inverse
distance weighting approach [@hermosilla2022]. The treed land cover
classes are broadleaf, coniferous, mixed wood, and wetland-treed, which
we use as a stratification. The land cover classification reached an
overall accuracy of 77.9% ± 4% [@hermosilla2022].

#### Forest Structure

Forest structural attributes (canopy height, canopy cover, structural
complexity \[coefficient of variation of height returns\], basal area,
aboveground biomass, and gross stem volume) for 2015 were extracted for
BC from the 30-m spatial resolution dataset generated by Matasci et al.
(2018a, 2018b). In brief, the method extends structural attributes
derived from a set of lidar acquisitions and field plots across Canada.
These data were generated for the treed pixels of Canada using the BAP
Landsat surface reflectance composite and auxiliary data (i.e.,
topography, geography) using a k-Nearest Neighbour imputation
[@matasci2018; @matasci2018b]. Accuracy metrics for imputed forest
structural attributes range from an RMSE of 24.5% (structural
complexity) to 82.3% (gross stem volume) and *R^2^* ranging from 0.125
(structural complexity) to 0.712 (gross stem volume) [@matasci2018;
@matasci2018b].

@@

The DHIs were calculated following
@razenkovaMediumresolutionDynamicHabitatInPress for the terrestrial area
of BC. In brief, we used Google Earth Engine [@gorelick2017] to obtain
valid Landsat pixels for the study area across a ten-year time span,
centered on 2015 (2011-2020), filtering out pixels containing shadows,
clouds, and cloud shadows using the QA band derived from the fmask
algorithm [@zhu2012], and then calculated the NDVI index. These
estimates were composited into a synthetic year of monthly data by
calculating the monthly median NDVI value. Finally, images of the sum,
minimum, and coefficient of variation across this synthetic year of NDVI
values were then calculated to represent the cumulative, minimum, and
variation DHIs, respectively. More detailed information can be found in
@razenkovaMediumresolutionDynamicHabitatInPress.

@@

## Sampling
